<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{QTL mapping using functional PCA for function-valued traits}
-->

# QTL mapping using functional PCA for function-valued traits


```{r knitr_options, echo=FALSE, results=FALSE}
library(knitr)
opts_chunk$set(fig.width = 10)
```

This vignette illustrates the use of the funqtl package for QTL
mapping with function-valued traits using functional principal
component analysis and multi-trait mapping.


## Data

```{r load_without_showing, echo=FALSE, results=FALSE}
library(funqtl)
data(simspal)
```

We will consider the analysis of a simulated data set, `simspal`.
There are `r nind(simspal)` recombinant inbred lines typed at a total
of `r totmar(simspal)` markers on `r nchr(simspal)` chromosomes.

We first load the funqtl package (which will also load R/qtl and other
packages) and the data.

```{r loading, eval=FALSE}
library(funqtl)
data(simspal)
```

Here's a quick summary of the data.

```{r summary_simspal}
summary(simspal)
```

The following plots the genetic marker map and the function-valued trait
for five of the RIL.

```{r plot_map_and_pheno}
par(mfrow=c(1,2))
plotMap(simspal, main="")
plot(1:241, simspal$pheno[160,], type="l", xlab="Time",
	  ylim=c(-120,0), ylab="Root Tip Angle (degrees)")
ind <- c(19, 20, 132, 72)
color <- c("blue", "red", "green", "orange")
for(i in seq(along=ind))
  lines(1:241, simspal$pheno[ind[i],], col=color[i])
```

There are `r nind(simspal)` individuals. The angle of the root tim is
measured for `r nphe(simspal)` time points, which are measured every 2
minutes for eight hours.


```{r plotlod}

out <- scanone(simspal, pheno.col=1:241, method="hk")
eff <- geteffects(simspal, pheno.cols=1:241)
nam <- phenames(simspal)
y <- as.numeric(substr(nam, 2, nchar(nam)))/60
plotlod(out, eff, y,  gap=15, horizontal = T)

```

This Figure shows the image of signed lod scores for each time points. The
x-axis represent time points and y-axis represent chromosome, lod
score represented as tone of the color. Red indicate positive effects
and blue color indicate negative effects.

Here, we've got lod score for all 241 time points. Our idea is to
reduce the dimension of 241 to fewer number.

1.  Get estimation of smooth function for each 162 individual to
   control measurment error.
    *  We need to decide how many base functions (b-spline) to be used
     estimating the smooth function.
    *  This step can be done using `cvfold()` function.
1.  Do functional PCA to these 162 individuals. Here we reduce the
   infinite dimensional function-valued data to small number of
   principal components.
1.  Do qtl mapping with these principal components. Combine the
   information using our proposed test statistics. (HKLOD, SL and ML)

```{r cvfold}
set.seed(1)
cvout <- cvfold(simspal, basisset = 15:60, fold = 10, random = F)
```

The 1st step is to decide how many basis functions to use estimating
our function-valued trait. If we use too much basis it will fit the
measurement error as well which is not meaningful. If we use too few
basis the estimation may not enough to represent the truth.

We've developed `cvfold` function to decide the number of basis
functions. `basisset` option
describe the number of basis sets to be tested. `fold` indicate how
many folders to be used in cross validation. In this code, we divide
our time points with equily spaced 10 sets. We used 9 folder to evaluate the
functional. And calculated sum of squared errors for the left one
folder.

The Figure shows the sum of error losses (y- axis) for each
number of basis from 15 to 60.

```{r cvfig, fig.height=8, fig.width = 10}
plot(15:60, cvout, xlab = "The number of basis",
     main = "sum of squared errors", type = "l")

```

```{r count}
which(cvout == min(cvout))

```

The minimum error loss calculated when we used 41 basis. So we decided
to use 41 B-spline basis functions.


## HKLOD, SL and ML scores


```{r hklodttc2}
cfpc <- calcfunpca(simspal, criteria=0.99, nbasis = 41)
Y <- cfpc$Y
eigfc <- cfpc$eigf

dim(Y)
Y[1:10,]
```

Now, we can used `calcfunpca` function to get small number of
principal components.

41 B-spline basis functions are used. the number of 4 PCs are selected
they explain more than 99 percent of data variation.

To convert those PCs back to original data information, we can use 'eigfc' object.

```{r backtoD}
# PC for 1st individual
Y[1,]

# plot the 1st individual ( raw data)
plot(1:241,simspal$pheno[1,], xlim=c(0,241), ylim=c(-90,0), ylab="angle", xlab="x")
par(new=T)
# plot the recovered data using PC
plot( -903.74*cfpc$eigf[1] + 222 *cfpc$eigf[2] -107*cfpc$eigf[3] -120 * cfpc$eigf[4], xlim=c(0,241), ylim=c(-90,0), col = "red", ylab="", xlab="")

```
The example shows how we can recover the 1st individual using PCs.



```{r scanoneM}

outhk <- scanoneM(simspal, Y = Y, method="hk")
outsl <- scanoneM(simspal, Y = Y, method="sl")
outml <- scanoneM(simspal, Y = Y, method="ml")

```

We proposed HKLOD, SL and ML scores, this statistic can be obtained
using `scanoneM` function.

```{r slodmlodcurve2, fig.height=10, fig.width = 10}
par(mfrow=c(3,1))
plot(outhk, ylim=c(0,8), main="The HKLOD curve for simspal data",
     bandcol="gray90")
abline(h=4.32, col="red", lty=3)
plot(outsl, ylim=c(0,2), main="The SL curve for simspal data",
     bandcol="gray90")
abline(h=1.05, col="red", lty=3)
plot(outml, ylim=c(0,4), main="The ML curve for simspal data",
     bandcol="gray90")
abline(h=3.05, col="red", lty=3)

# permutation threshold
# permouthk <- scanoneF(simspal, Y=Y, method = "hk", n.perm=1000)
# permoutf <- scanoneF(simspal, Y=Y, method = "f", n.perm=1000)
# permoutsl <- scanoneF(simspal, Y=Y, method = "sl", n.perm=1000)
# permoutml <- scanoneF(simspal, Y=Y, method = "ml", n.perm=1000)
# summary(permouthk) # display 5, 10 % threshold
# summary(permoutf) # display 5, 10 % threshold
# summary(permoutsl) # display 5, 10 % threshold
# summary(permoutml) # display 5, 10 % threshold
```

The results are in Figure above. Horizontal lines indicate the 5%
genome-wide significance thresholds, derived by a permutation test. We
didn't run the code (for permutation test) above since it takes a long time (about one hour maybe) .

## Getting multiple QTL

```{r stepwiseqtlM}
qtlhk <- stepwiseqtlM(simspal, Y = Y, max.qtl=6 , method = "hk", penalties = c(4.44, 10, 18), additive.only = T )

qtlml <- stepwiseqtlM(simspal, Y = Y, max.qtl=8 , method = "ml", penalties = c(2.05, 8, 8), additive.only = T )

qtlsl <- stepwiseqtlM(simspal, Y = Y, max.qtl=8 , method = "sl", penalties = c(1.07, 8, 8), additive.only = T )

qtlhk
qtlsl
qtlml
```

`stepwiseqtlM` function returns qtl object by using stepwise QTL
selection.

``` {r profilecode, fig.height = 8, fig.width = 10}
par(mfrow=c(3,1))

    rqtlhk <- refineqtlM(cross = simspal, Y = Y, qtl = qtlhk, method = "hk" )
    rqtlsl <- refineqtlM(cross = simspal, Y = Y, qtl = qtlsl, method = "sl" )
    rqtlml <- refineqtlM(cross = simspal, Y = Y, qtl = qtlml, method = "ml" )

plotLodProfile(rqtlhk, ylab = "Profile LOD score", main = "HKLOD Profile")
plotLodProfile(rqtlsl, ylab = "Profile LOD score", main = "SL Profile")
plotLodProfile(rqtlml, ylab = "Profile LOD score", main = "ML Profile")
```

You can also make profile curves by using `refineqtlM` and
`plotLodProfile` function. Here are HKLOD, SL and ML profile curves. All
the method found 2-QTL model.

## Effect plot

```{r effectplot2}

## getting Principal componetes.
Z <- calcfunpca(simspal, n.max=8, criteria = .99, nbasis = 30)

### This can be manualy done as follows ( this step needed to get eigen functions)

Y = t(simspal$pheno)
m = 241 # time dimension

## creating 30 b-spline basis using 4 knots
splinebasis.y <- create.bspline.basis(c(0, m), 30, 4)
time <- 0:(m - 1) + 0.5

## do smoothing
mat <- eval.basis(time, splinebasis.y)
coef.y <- solve(crossprod(mat), crossprod(mat, Y))

## get smoothed Y phenotype. now Y is n number of functionals
yfd = fd(coef.y, splinebasis.y, list("time", "indv", "value"))

## Do functional PCA
y.pcalist3 = pca.fd(yfd, 20)

## The proportion variance explained from eigen functions
y.pcalist3$varprop

## first 4 eigen values explain more than 99 percent of data variation.
sum(y.pcalist3$varprop[1:4])
sum(y.pcalist3$varprop[1:3])


y.pcalist3 = pca.fd(yfd, 4)
eigfc3 <- y.pcalist3$harmonics
mat3 <- eval.fd(time, eigfc3)
nY3 <- t(solve(crossprod(mat3), crossprod(mat3, Y)))

temp <- simspal
temp$pheno[,1] <- Z$Y[,1]
temp$pheno[,2] <- Z$Y[,2]
temp$pheno[,3] <- Z$Y[,3]
temp$pheno[,4] <- Z$Y[,4]

hksleff <- vector("list", 4)
for(i in 1:4) {
  hksleff[[i]] <- summary(fitqtl(temp, phe=i, qtl=rqtlhk, method="hk", get.ests=TRUE, dropone=FALSE))$ests[,1]*c(1,2,2)
}

nam <- names(hksleff[[1]])
hksleff <- matrix(unlist(hksleff), byrow=TRUE, ncol=length(nam))
colnames(hksleff) <- nam

```
By using these 4 eigen functions, we can recover effect functions.
`hksleff` have the coefficients for baseline, Q1 and Q2 effect
functions. We can draw baseline, Q1 and Q2 effect function as follows.

```{r effplot3, fig.height=5, fig.width = 10}
par(mfrow=c(1,3))
par(las=1)

baselinehk <- hksleff[1,1] * eigfc3[1]
for(j in 2:4)
    baselinehk = baselinehk + hksleff[j,1] * eigfc3[j]

time <- (0:240)/30
x = (0:240)

plot(baselinehk, xlim = c(0,250), ylim=c(-110,0), xaxt = "n", ylab="", col="red")
axis(1, at = x[c(1,61,121,181,241)] , labels = time[c(1,61,121,181,241)])
mtext("Time (hours)", side=1, line = 2.8, cex = .7)
mtext("Tip angle (degrees)", side=2, line = 2.8, las=3, cex = .7)

q1hk <- hksleff[1,2] * eigfc3[1]
for(j in 2:4)
    q1hk = q1hk + hksleff[j,2] * eigfc3[j]

plot(q1hk, xlim = c(0,250), ylim=c(-5,9), xaxt = "n", col="red", ylab="")
axis(1, at = x[c(1,61,121,181,241)] , labels = time[c(1,61,121,181,241)])
mtext("Time (hours)", side=1, line = 2.8, cex = .7)
mtext("QTL effect", side=2, line = 2.8, las=3, cex = .7)
abline(h = 0, lty=2)

q2hk <- hksleff[1,3] * eigfc3[1]
for(j in 2:4)
    q2hk = q2hk + hksleff[j,3] * eigfc3[j]

plot(q2hk, xlim = c(0,250), ylim=c(-5,9), xaxt = "n", col="red", ylab="")
axis(1, at = x[c(1,61,121,181,241)] , labels = time[c(1,61,121,181,241)])
mtext("Time (hours)", side=1, line = 2.8, cex = .7)
mtext("QTL effect", side=2, line = 2.8, las=3, cex = .7)
abline(h = 0, lty=2)
```

